Staged Abstract Interpreters

### 2019-09-06 paper/50_stageabs.tex
revision
typo

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter, now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section has the same abstract semantics with the
unstaged version from Section~\ref{unstaged_abs}.  The approach that refactors
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore has no
need to rewrite the analysis. We first present the staged lattices and staged
monads, and then the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. At last, we discuss several optimizations.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve stage polymorphism. Now we instantiate the type @R@ to @Rep@ and
still use power-set as an example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is obtained by specializing @Lattice@ with @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
are eventually delegated to operations on a staged set data type, such as
@subsetOf@, @union@ and @intersect@. We even do not have to change the
implementation code, except the types -- from @Set[T]@ to @Rep[Set[T]]@, as shown
in light gray code. The LMS library provides implicit conversions that lift
current-stage values to next-stage values. In the code generation part, these
operations emit their corresponding next-stage code. Again, other lattice
structures such as products and maps are implemented in a similar way.

@@

Now we can implement the staged abstract semantics, where @R@ is instantiated as
@Rep@, and the types of abstract components are reused from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. We manually squeeze the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effect and @B@ for the state effect.
In our monad stack scheme, they are both instantiated with type @Cache@. Now the
grounded result type is a pair of two staged value: @Rep[Set[(Value, Store)]]@ and
@Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations} When deriving the staged concrete interpreter
from the unstaged counterpart, we notice that the recursive calls of @ev@ on the
lambda body are shifted from @ap_clo@ to @close@. In other words, we need to
eagerly specialize the interpreter and generate code every time when we see a
$\lambda$ term, instead of lazily when applications happen.
With no surprise, it is similar in the staged abstract interpreter;
additionally, we have to handle nondeterminism occurred by over-approximation
of runtime behaviors.
In the rest of this part, we keep our roadmap and focus on discussing closure
representations and function applications for the staged version.

The @close@ method now is a mixing of the \textit{staged} concrete version and
unstaged \textit{abstract} version: we build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the @Ans@ monad to
grounded values of type @Result@ by providing the desired arguments, i.e., the
new environment, new store, and caches. The collapsing of monad happens at
current-stage, so the call of @ev@ is unfolded through staging. Finally, we
generate a singleton set containing the compiled next-stage closure
@emit_compiled_clo(f)@, represented by an IR node in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure @clo@ in the set, by using @emit_ap_clo@. The
method @emit_ap_clo@ generates the current-stage representation of the
future-stage result of application.
Finally, we put the future @out@ cache, store, and values back into
the current-stage monads.
\begin{lstlisting}
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](rator)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

\paragraph{Staged Caching and Fixpoint Iteration} Our fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @out@ cache contains the current
configuration will produce a next-stage Boolean value, i.e., of type
@Rep[Boolean]@. Consequently, the branching condition cannot be determined
statically -- we need to generate code for @if@. Figure
\ref{fig:staged_coind_cache} shows the staged version of @fix@. The
monadic-binded variable @res@ is a next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of queried result form
the @out@ cache and the @out@ cache itself. The else branch constructs a monadic
value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@ cache.
After which, we use the similar technique that collapses the monad @m@ to
grounded values by providing its desired environment and etc. Finally, we have a
current-stage representation of the future-stage values @res@, and put the
content of @res@ back into the monad stack.

\begin{figure}[h!]
\centering
\begin{lstlisting}
  def fix_cache(e: Expr): Ans = for {
    ρ   <- ask_env;  σ <- get_store;  in <- ask_in_cache;  out <- get_out_cache
    cfg <- lift_nd[Config](Set((unit(e), ρ, σ)))
    res <- lift_nd[(Set[ValSt], Cache)](Set(
      // a next-stage value of type Rep[(Set[ValSt], Cache)] generated by if
      if (out.contains(cfg)) (repMapToMapOps(out).apply(cfg).toList, out)
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- eval(fix_cache)(e);  σ <- get_store;  _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }))
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\vspace{-2em}
\end{figure}

\subsection{A Little Bit of Code Generation, Again}
The code generation for compiled closures and function applications are similar
to the staged concrete interpreter. We have two the IR node as case classes, they
take additional caches as arguments. We also elide how to generate next-stage Scala
code for them.
\begin{lstlisting}
case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                          => Rep[(Set[ValSt], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                    in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

@@

\iffalse
Revision: Solving Practical Challenges.
theoretically, all the things should work nicely. Unfolding the interpreter over the AST.
But the generated code is blowed up. For example .... This should not affect the correctness,
but poses burden on the MSP system (ie LMS) and the next stage compiler/runtime (ie, scalac and JVM).
1) LMS becomes slower since a large IR graph is contructed during the staging.
2) Scalac becomes slower when reading a such large source code.
JVM has certain limitation on the size of a single method.

TODO: can we formulate selective caching as a partially-static data law.
TODO: lambda lifting for if
\fi

@@

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistence, i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal Form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularly.
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structures we used here are simply
    black boxes that wrap the Scala's library data structures (in the next
    stage).
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}

>>>>

In the previous sections, we have seen an unstaged abstract interpreter and a
staged concrete interpreter. Now we begin describing the implementation of
their confluence -- a staged abstract interpreter.  Unsurprisingly, the staged
abstract interpreter in this section uses the same abstract semantics from the
unstaged version in Section~\ref{unstaged_abs}.  The approach we refactor
the unstaged one to the staged abstract interpreter is modular, and does not
sacrifice soundness or precision. The designer of the analyzer therefore has no
need to rewrite the analysis. We first present the staged lattices and staged
monads, and then discuss the staged version of primitive operations, especially
@close@, @ap_clo@ and @fix@. At last, we discuss several optimizations.

@@

In Section \ref{stagedpoly_lat}, we exploited the higher-kinded type @R@ to
achieve the stage polymorphism. Now we instantiate the type @R@ as @Rep@ and
still use power-set as example to describe its staged version.
\begin{lstlisting}[escapechar=!]
  trait RepLattice[S] extends Lattice[S, Rep]
  def RepSetLattice[T]: RepLattice[Set[T]] = new RepLattice[Set[T]] {
    lazy val ⊥: !\hl{Rep[Set[T]]}! = Set[T]()
    lazy val ⊤: !\hl{Rep[Set[T]]}! = error("No representation for ⊤")
    def ⊑(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Boolean]}! = l1 subsetOf  l2
    def ⊔(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 union     l2
    def ⊓(l1: !\hl{Rep[Set[T]]}!, l2: !\hl{Rep[Set[T]]}!): !\hl{Rep[Set[T]]}!  = l1 intersect l2
  }
\end{lstlisting}

The trait @RepLattice@ is shown by specializing @Lattice@ with type @Rep@.
The @RepSetLattice@ is an instance of @RepLattice@, where the lattice operations
are eventually delegated to the operations on a staged set data type, such as
@subsetOf@, @union@ and @intersect@. We do not need to change the
implementation code, except the types -- such as from @Set[T]@ to @Rep[Set[T]]@.
We show the changes in light gray code. The LMS library provides implicit
conversions that lift the current-stage values to their next-stage values. In
the code generation part, these operations emit their corresponding next-stage
code. Again, other lattice structures such as products and maps are implemented
in a similar way.

@@

Now we can implement the staged abstract semantics, under that @R@ is instantiated as
@Rep@. The types of abstract components are reused from the unstaged version.

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure from the unstaged abstract interpreter,
but replacing \textit{all} the monad transformers to their staged version. The
following code snippet shows this change. We manually fuse the three inner
transformers (@IdM@ omitted) into a single monad @RepSetReaderStateM[A, B, C]@,
where @A@ is the type parameter for the reader effects and @B@ for the state effects.
In our monad stack scheme, they are both instantiated with the type @Cache@. 
Similar to the unstaged version, the grounded result type is a pair of two
staged value: @Rep[Set[(Value, Store)]]@ and @Rep[Cache]@.
\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type AnsM[T] = RepReaderT[RepStateT[RepSetReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[Set[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\paragraph{Primitive Operations} 
When deriving the staged concrete interpreter from its unstaged counterpart, we
notice that the recursive calls of the evaluator @ev@ on the lambda body are shifted from
@ap_clo@ to @close@. In other words, we eagerly specialize the interpreter and
generate code every time when we see a $\lambda$ term, instead of lazily when the
applications happen.  With no surprise, it is similar here. Additionally,
in the staged abstract interpreter, we have to handle nondeterminism
incurred by the over-approximation of runtime behavior.
In the rest of this part, we keep our roadmap and focus on discussing closure
representations and function applications for the staged version.

The @close@ method now is a mix of the \textit{staged} concrete version and
unstaged \textit{abstract} version. We first build a current-stage function @f@, which
takes four next-stage values, including the argument and latest store, and
@in/out@ caches additionally. Inside of @f@, we collapse the @Ans@ monadic value to
a grounded value of type @Result@, by providing the desired arguments, i.e., the
new environment, new store, and caches. The collapsing of monad happens at
current-stage, so the calls of @ev@ is unfolded within staging. Finally, we
generate a singleton set containing the compiled next-stage closure
@emit_compiled_clo(f)@, which is represented by an IR node in LMS.
\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(Set[(Value,Store)], Cache)], λ: Lam, ρ: Exp[Env]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(Set[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x)
        ev(e)(ρ + (unit(x) → α))(σ ⊔ Map(α → arg))(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f, λ, ρ))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the next-stage set of closures into the monad stack, and generate a
next-stage value for each closure @clo@ in the set, by using @emit_ap_clo@. The
method @emit_ap_clo@ generates the current-stage representation of the
future-stage application result.
Finally, we put the future @out@ cache, store, and values back into
the current-stage monads.
\begin{lstlisting}
  def emit_ap_clo(fun: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(Set[ValSt], Cache)]
  def ap_clo(ev: Expr => Ans)(funs: Rep[Value], arg: Rep[Value]): Ans = for {
    σ   <- get_store;      clo <- lift_nd[AbsValue](funs)
    in  <- ask_in_cache;   out <- get_out_cache
    res <- lift_nd[(Set[ValSt], Cache)](Set(emit_ap_clo(clo, arg, σ, in, out)))
    _   <- put_out_cache(res._2);  vs <- lift_nd[ValSt](res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

\paragraph{Staged Caching and Fixpoint Iteration} Our fixed-point iteration
again relies on the two caches @in@ and @out@, which are both staged maps now.
Therefore, the query of whether the @in/out@ cache contains the current
configuration will produce next-stage Boolean values, i.e., of type
@Rep[Boolean]@. Consequently, the branching condition cannot be determined
statically -- we need to generate code for the whole @if@ expression. Figure
\ref{fig:staged_coind_cache} shows the staged version of @fix@. The
variable @res@ represents the next-stage result, consisting of a next-stage
@if@ expression. The true branch simply returns a pair of the queried result form
the @out@ cache and the @out@ cache itself. The else branch constructs a monadic
value @m@ of type @Ans@ first, which evaluates @e@ under the new @out@ cache.
After which, we use a similar technique that collapses the monadic value @m@ to
grounded values, by providing its desired environment and etc. Finally, we have a
current-stage representation of the future-stage values @res@, and put the
content of @res@ back into the monad stack.

\begin{figure}[h!]
\centering
\begin{lstlisting}
  def fix_cache(e: Expr): Ans = for {
    ρ   <- ask_env;  σ <- get_store;  in <- ask_in_cache;  out <- get_out_cache
    cfg <- lift_nd[Config](Set((unit(e), ρ, σ)))
    res <- lift_nd[(Set[ValSt], Cache)](Set(
      // a next-stage value of type Rep[(Set[ValSt], Cache)] generated by if
      if (out.contains(cfg)) (out(cfg), out)
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- eval(fix_cache)(e);  σ <- get_store;  _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }))
    _ <- put_out_cache(res._2);  vs <- lift_nd(res._1);  _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\vspace{-2em}
\end{figure}

\subsection{A Little Bit of Code Generation, Again}
The code generation for compiled closures and function applications are similar
to their counterparts in the staged concrete interpreter. We have two IR nodes
implemented as case classes; they also take additional caches as arguments. We
elide the code generation part for these IR nodes.
\begin{lstlisting}
case class IRCompiledClo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                          => Rep[(Set[ValSt], Cache)], λ: Lam, ρ: Rep[Env]) extends Def[AbsValue]
case class IRApClo(clo: Rep[AbsValue], arg: Rep[Value], σ: Rep[Store],
                    in: Rep[Cache], out: Rep[Cache]) extends Def[(Set[(Value, Store)], Cache)]
\end{lstlisting}

@@

@@

\begin{itemize}
  \item We usually assume the underlying MSP system achieves cross-stage
    persistence, i.e., the generated program does not change the observational
    behaviors of the original program. In the case of this paper, we use the
    LMS framework, which 1) relies on the type checking facilities of Scala, and
    2) generates code in A-Normal Form \cite{Flanagan:1993:ECC:155090.155113}
    that preserves the execution order within a stage \cite{DBLP:conf/birthday/Rompf16}.
    However, in general, it is possible to subvert these guarantees by
    using unsafe features, such as casts.
  \item As shown in the previous section, the generic interpreter is untouched
    and shared by the four artifacts. We only instantiate binding-time
    annotations and the monadic types. This allows the programmer to easily
    check the correctness of implementation of staged monads modularly.
  \item We build the staged monads and staged data structures in a
    correct-by-construct way that directly corresponds to their unstaged
    version. For instance, the staged data structures we used here are simply
    black boxes that wrap the Scala's library data structures.
  \item In our experiments, the staged abstract interpreter produces the same
    result with the unstaged one on all benchmark programs we tested.
\end{itemize}
